---
title: "main"
author: "Frank Edwards"
date: "6/18/21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(tidyverse)

theme_set(theme_minimal())


## read in fatal encounters as fe
## normalize for time covered in dataset

## time constant to normalize to years
t_norm<-as.numeric(max(fe$date) - min(fe$date))/365.25

source("read_fe.R")

## read in 14-18 ACS nhgis pop data as
## aianh_dat; county_dat; state_dat; nation_dat; aian_shp_file

source("read_pop.r")

### read and join shape files
source("join_shapes.r")

### compute distance between deaths and aianh borders / ms river
source("distance_to_aianh.r")


### make pl280 variable
make_pl280<-function(x){
  x$pl280<-case_when(
    x$state %in% c("CA", "MN", "NE", "OR", 
                   "WI", "AK") ~ "Mandatory",
    x$state %in% c("NV", "FL", "ID", "IA", 
                   "WA", "SD", "MT", "ND", 
                   "AZ", "UT") ~ "Optional",
    T ~ "Non-PL280"
  )
  return(x)
}


fe<-make_pl280(fe)

## state-level risk
state_counts<-fe %>% 
  rename(sex = gender) %>% 
  group_by(state, sex, age_cat) %>% 
  summarise(n = n()) %>%
  full_join(age_sex_pop %>% 
              rename(age_cat = age) %>% 
              filter(race_ethn == "aian") %>% 
              group_by(state, sex, age_cat) %>% 
              summarise(pop = sum(pop))) %>% 
  mutate(n = ifelse(is.na(n), 0, n)) ## fill in zero deaths

### national risks

natl_counts<-fe %>% 
  rename(sex = gender) %>% 
  group_by(sex, age_cat) %>% 
  summarise(n = n()) %>%
  full_join(age_sex_pop %>% 
               rename(age_cat = age) %>% 
               filter(race_ethn == "aian") %>% 
               group_by(sex, age_cat) %>% 
              summarise(pop = sum(pop))) %>% 
  mutate(n = ifelse(is.na(n), 0, n))   ## fill in zero deaths

### compute age-specific death rates for state and natl
natl_counts<-natl_counts %>% 
  mutate(asdr = n / pop / t_norm)

state_counts<-state_counts %>% 
  mutate(asdr = n / pop / t_norm)

### compute age-adjusted death rate for states
state_adr<-state_counts %>% 
  left_join(natl_counts %>% 
              select(sex, age_cat, pop) %>% 
              rename(pop_natl = pop)) %>% 
  mutate(adr_count = asdr * pop_natl) %>% 
  group_by(state, sex) %>% 
  summarise(n = sum(n, na.rm=T), 
            adr_count = sum(adr_count, na.rm=T), 
            pop = sum(pop, na.rm=T), 
            pop_natl = sum(pop_natl, na.rm=T)) %>% 
  mutate(adr = adr_count / pop_natl * 1e5,
         cdr = n / pop / t_norm * 1e5)

## compute age-adjusted death rate for census divisions
### attach census division from 2010 census file
div_counts<-state_counts %>% 
  left_join(state_dat %>% 
              select(state, DIVISIONA)) %>% 
  group_by(DIVISIONA, sex, age_cat) %>% 
  summarise(n=sum(n), pop = sum(pop)) %>% 
  ungroup() %>% 
  mutate(asdr = n / pop / t_norm) %>% 
  left_join(natl_counts %>% 
              select(sex, age_cat, pop) %>% 
              rename(pop_natl = pop)) %>% 
  mutate(adr_count = asdr * pop_natl) %>% 
  group_by(DIVISIONA, sex) %>% 
  summarise(n = sum(n), adr_count = sum(adr_count), pop = sum(pop), pop_natl = sum(pop_natl)) %>% 
  mutate(adr = adr_count / pop_natl * 1e5,
         cdr = n / pop / t_norm * 1e5)  


#### VISUALS FOR ADR, ASDR

### Age spec visual
### convert age to numeric for plotting
age_recode<-function(df){
  age_recode<-df%>%
    mutate(age = 
      case_when(
        age=="0" ~ "0",
        age=="1-4" ~ "1",
        age=="5-9" ~ "5",
        age=="85+" ~ "85",
        !age%in%c("0", "85+") ~ substr(age, 1, 2)
      ),
      age = as.numeric(age))
      return(age_recode)
}

natl_counts<-age_recode(natl_counts %>% 
                          rename(age = age_cat))

ggplot(natl_counts,
       aes(x = age, y = asdr * 1e5)) + 
  geom_line() + 
  facet_wrap(~sex, scales = "free") + 
  labs(title = "AIAN age-specific death rates per 100,000",
       subtitle = "US 2011 - 2020",
       x = "Age",
       y = "Rate") + 
  ggsave("./vis/age_spec_nat.png", width = 7)

### State ADR 

library(usmap)
us_map <- usmap::us_map() 

us_map<-us_map %>% 
  rename(state = abbr)

state_adr_map<-state_adr %>% 
  right_join(us_map) 

library(viridisLite)
library(gridExtra)

p1<-ggplot(state_adr_map %>% 
             filter(sex == "Male"),
       aes(x=x, y = y, fill = adr,
           group = group)) + 
  geom_polygon(color = "black") + 
  theme_void() +
  labs(fill = "") +
  theme(legend.position = "bottom") + 
  scale_fill_viridis_c() + 
  labs(subtitle = "Male age-adjusted mortality")

p2<-ggplot(state_adr_map %>% 
             filter(sex == "Female"),
       aes(x=x, y = y, fill = adr,
           group = group)) + 
  geom_polygon(color = "black") + 
  theme_void() +
  labs(fill = "") +
  theme(legend.position = "bottom") + 
  scale_fill_viridis_c() + 
  labs(subtitle = "Female age-adjusted mortality")

g<-arrangeGrob(p1, p2)
ggsave(file = "./vis/adr_state.png", g, width = 8, height = 12)

### Division ADR
div_map<-state_adr_map %>% 
  select(-n, -adr_count, -pop, -pop_natl, -adr, -cdr) %>% 
  left_join(state_dat %>% 
              select(state, DIVISIONA)) %>% 
  left_join(div_counts) 
  
p1<-ggplot(div_map %>% 
             filter(sex == "Male"),
       aes(x=x, y = y, fill = adr,
           group = group)) + 
  geom_polygon(color = "black") + 
  theme_void() +
  labs(fill = "") +
  theme(legend.position = "bottom") + 
  scale_fill_viridis_c() + 
  labs(subtitle = "Male age-adjusted mortality")

p2<-ggplot(div_map %>% 
             filter(sex == "Female"),
       aes(x=x, y = y, fill = adr,
           group = group)) + 
  geom_polygon(color = "black") + 
  theme_void() +
  labs(fill = "") +
  theme(legend.position = "bottom") + 
  scale_fill_viridis_c() + 
  labs(subtitle = "Female age-adjusted mortality")

g<-arrangeGrob(p1, p2)
ggsave(file = "./vis/adr_division.png", g, width = 8, height = 12)

### format state-level data


```



## Descriptives




## H1: Native death by law enforcement will become more common (or increase) as one moves west of the Mississippi River.

```{r}
### n by west and east
table(fe_sf$distance_ms<0)

sum(fe_sf$distance_ms<0)/nrow(fe_sf)
### 94 % west of MS

ggplot(fe_sf) + 
  geom_histogram(aes(x = distance_ms)) + 
  geom_vline(aes(xintercept = 0), lty = 2) +
  labs(title = "Figure 1. Histogram of location of death's proximity to MS River",
       subtitle = "negative numbers are west, positive numbers are east")
```

Figure 1 shows the distribution of AIAN people killed by police documented in Fatal Encounters relative to the nearest point on the Mississippi River. Deaths are far more common West of the MS river, but do not increase consistently as one moves west. They peak within 500 km of the MS River, with a local maximum about 2000 km from the MS river. ADD MORE DETAIL

## H2: Native death by law enforcement will be greatest at high-contact zones such as reservation border towns, and urban centers with a high population of Native Peoples.
 
```{r}
ggplot(fe_sf) + 
  geom_histogram(aes(x = distance_aian)) + 
  labs(title = "Figure 2. Proximity to Native land when killed",
       subtitle = "Distance of zero indicates on native land")
```

```{r}
proximity_table<-fe_sf %>% 
  mutate(location = case_when(
    distance_aian == 0 ~ "On AIAN land",
    distance_aian <= 10 ~ "Within 10km of AIAN border",
    distance_aian <= 50 ~"Within 50km of AIAN border",
    T ~ "Further than 50km from AIAN border"
  )) %>% 
  group_by(location) %>% 
  summarise(n = n())

st_geometry(proximity_table)<-NULL

knitr::kable(proximity_table,
             caption = "Table 1. Proximity to Native land when killed by police")
```

As shown in figure 2 and table 1, more than 80 percent of people killed by police identified as AIAN in Fatal Encounters were killed either on or within 50km of Native lands. MORE ON THIS LATER

## H3: Native death by law enforcement will occur more often in geographic locations most impacted by Termination Era federal Indian policy (1947-1961)


```{r}
## by agency type

### make agency type deaths per 100,000 per year for national pop
### make pl280 population scaled for pl280 states

counts_280<-make_pl280(state_counts) %>% 
  group_by(pl280, age_cat, sex) %>% 
  summarise(pop = sum(pop),
            n = sum(n)) %>% 
  rename(age = age_cat) %>% 
  age_recode()

### compute age-specific death rates for state and natl
counts_280<-counts_280 %>% 
  mutate(asdr = n / pop / t_norm)

### compute age-specific death rates for state and natl
### compute age-adjusted death rate for states
adr_280<-counts_280 %>% 
  left_join(natl_counts %>% 
              select(sex, age, pop) %>% 
              rename(pop_natl = pop)) %>% 
  mutate(adr_count = asdr * pop_natl) %>% 
  group_by(pl280, sex) %>% 
  summarise(n = sum(n, na.rm=T), 
            adr_count = sum(adr_count, na.rm=T), 
            pop = sum(pop, na.rm=T), 
            pop_natl = sum(pop_natl, na.rm=T)) %>% 
  mutate(adr = adr_count / pop_natl * 1e5,
         cdr = n / pop / t_norm * 1e5)

### death rates by pl280 state type
ggplot(adr_280,
       aes(x = sex, y = adr)) + 
  geom_col() + 
  coord_flip() + 
  facet_wrap(~pl280)


agency_type<-fe %>% 
  group_by(agency_type) %>% 
  summarise(n=n()) 

ggplot(agency_type,
       aes(x = n, y = reorder(agency_type, n))) +
  geom_col() + 
  labs(x = "AIAN people killed by police 2011 - 2020",
       y = "Law enforcement agency type")

```



THIS ALL WORKS

NOW DO CORRELATION WITH FOSTER CARE AND MAPS

